---
title: "surveyframe: {{DEMO_TITLE}}"
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(surveyframe)

`%||%` <- function(x, y) if (is.null(x)) y else x

show_value <- function(x) {
  if (is.null(x) || length(x) == 0) return("")
  if (is.list(x)) {
    return(paste(sprintf("%s: %s", names(x),
                         vapply(x, show_value, character(1))), collapse = "; "))
  }
  paste(as.character(x), collapse = ", ")
}

plan_table <- function(instrument) {
  plan <- sf_plan(instrument)
  if (!length(plan)) return(data.frame())
  do.call(rbind, lapply(seq_along(plan), function(i) {
    b <- plan[[i]]
    data.frame(
      order = i,
      id = b$id,
      question = b$research_question,
      family = b$family %||% "",
      method = b$method %||% b$test %||% "",
      roles = show_value(b$roles %||% b$variables),
      options = show_value(b$options),
      decision_rule = b$decision_rule %||% b$interpretation %||% "",
      needs_data = !isFALSE(b$requires_data),
      stringsAsFactors = FALSE,
      check.names = FALSE
    )
  }))
}
```

## The study

This notebook runs the `{{DEMO_NAME}}` demo end to end: load the instrument,
read the responses, run the pre-declared analysis plan, and render a report.
It is yours to change. Replace the demo with your own instrument and your own
data and the rest of the notebook still works.

```{r load}
demo       <- sframe_demo("{{DEMO_NAME}}")
instrument <- demo$instrument
responses  <- demo$responses
instrument
```

`{{DEMO_TEACHES}}`

The demo contains four auditable artifacts. Presentation and provenance demos
may reuse the response data from their matching analysis demo.

```{r artifacts}
data.frame(
  artifact = c("Instrument", "Responses", "Codebook", "Expected results"),
  file = basename(unlist(demo[c(
    "instrument_path", "responses_path", "codebook_path", "results_path"
  )])),
  stringsAsFactors = FALSE
) |>
  knitr::kable()
```

## The questions

```{r items}
codebook <- as.data.frame(codebook_report(instrument))
knitr::kable(codebook[, c("id", "label", "type")])
```

## The plan, declared before the data existed

This is what makes the analysis a contract rather than a search. The blocks
below were written into the instrument at design time.

```{r plan}
knitr::kable(plan_table(instrument))
```

The first column is the run and report sequence. `run_analysis_plan()` executes
the plan from top to bottom, and the report keeps the same order. In
SurveyBuilder, drag the plan cards into their final order on the Plan or Run
Preview tab before examining results.

The equivalent R operation is shown below but is deliberately not run here.
Reorder by stable block ID, validate the result, and save it before data
collection. A change after collection begins should retain the earlier
instrument and be disclosed with `amend_sframe()`.

```{r reorder, eval = FALSE}
plan <- sf_plan(instrument)
ids <- vapply(plan, `[[`, character(1), "id")
wanted_order <- rev(ids)             # replace with the order you planned
stopifnot(setequal(wanted_order, ids))
sf_plan(instrument) <- plan[match(wanted_order, ids)]
instrument <- as_sframe(validate_sframe(instrument, strict = TRUE))
```

## Validate and screen the inputs

`sframe_demo()` has already read the CSV through `read_responses()`. With your
own files, perform that import explicitly so unexpected columns or values are
caught before analysis.

```{r validate}
validation <- validate_sframe(instrument, strict = FALSE)
validation

dim(responses)
head(responses, 3)
```

```{r real-import, eval = FALSE}
responses <- read_responses(
  demo$responses_path,
  instrument,
  respondent_id = "respondent_id",
  meta_cols = c("started_at", "submitted_at"),
  strict = TRUE
)
```

Screen missingness and response quality before interpreting an analysis. Some
demos contain no scales or attention checks. Read an empty or
not-applicable section as information about the demo, not a reason to skip
the step.

```{r screen}
missing <- missing_data_report(responses, instrument)
knitr::kable(as.data.frame(missing))

quality <- quality_report(responses, instrument)
knitr::kable(as.data.frame(quality))
```

```{r reliability}
if (length(sf_scales(instrument)) > 0 &&
    requireNamespace("psych", quietly = TRUE)) {
  reliability <- reliability_report(
    responses, instrument,
    omega = TRUE
  )
  knitr::kable(as.data.frame(reliability))
} else if (length(sf_scales(instrument)) == 0) {
  cat("This instrument has no scored scales, so reliability is not applicable.\n")
} else {
  cat("Install the optional psych package to compute reliability.\n")
}

## Run the plan once

```{r run}
results <- run_analysis_plan(
  responses,
  instrument,
  plots = requireNamespace("ggplot2", quietly = TRUE),
  seed = 20260828L,
  strict = FALSE
)
status <- attr(results, "status")
status
attr(results, "seed")
knitr::kable(as.data.frame(results))
```

A normal non-strict return may contain failed blocks, for example when an
optional method package is unavailable, so always inspect `status`. For a
verification or release run, rerun with `strict = TRUE` so any failed block
stops the workflow.

```{r strict-run, eval = FALSE}
results <- run_analysis_plan(
  responses, instrument, plots = TRUE, seed = 20260828L, strict = TRUE
)
```

## Read the complete results

Each result stays paired with the plan block that produced it. The display
below keeps the planned method and decision rule beside the statistic, table,
plot, reproducible syntax where supported, interpretation prompt, and
references.

```{r result-details, results = "asis"}
plan <- sf_plan(instrument)
for (b in results) {
  planned <- plan[[match(b$block_id, vapply(plan, `[[`, character(1), "id"))]]
  cat("\n###", b$block_id, b$research_question, "\n\n")
  cat("**Planned method:**", planned$method %||% planned$test, "\n\n")
  cat("**Roles:**", show_value(planned$roles %||% planned$variables), "\n\n")
  if (nzchar(planned$decision_rule %||% planned$interpretation %||% "")) {
    cat("**Planned decision rule:**",
        planned$decision_rule %||% planned$interpretation, "\n\n")
  }
  if (!is.null(b$error)) {
    cat("**This analysis failed.** ", b$error, "\n\n", sep = "")
    next
  }
  if (!is.null(b$apa)) cat(b$apa, "\n\n")
  if (!is.null(b$effect_label)) cat("**Effect:**", b$effect_label, "\n\n")
  if (is.data.frame(b$table)) print(knitr::kable(b$table))
  if (!is.null(b$syntax)) cat("\n```r\n", b$syntax, "\n```\n", sep = "")
  if (!is.null(b$plot)) print(b$plot)
  if (length(b$diagnostic_plots %||% list())) {
    for (p in b$diagnostic_plots) print(p)
  }
  resolved <- analysis_syntax(b)
  if (length(resolved)) cat("\n**Resolved R call**\n\n```r\n",
                            paste(resolved, collapse = "\n"), "\n```\n", sep = "")
  if (nzchar(b$prompt %||% "")) cat("\n**Interpretation prompt:**", b$prompt, "\n\n")
  refs <- unlist(b$citations %||% character(0), use.names = FALSE)
  if (length(refs)) cat("**References:**", paste(refs, collapse = "; "), "\n\n")
}
if (!is.null(status) && status$failed > 0) {
  cat("\n", status$failed, " of ", status$blocks, " analyses failed: ",
      paste(status$failed_blocks, collapse = ", "), ".\n", sep = "")
}
```

## Check the bundled expected results

The expected-results artifact is a compact, long-format reference generated
for this demo. Display it beside the live run when checking another package or
investigating a changed number. Numerical comparison needs method-appropriate
tolerances, and compare the numeric values, not formatted strings, since not
every quantity is exact.

```{r expected}
expected <- utils::read.csv(demo$results_path, stringsAsFactors = FALSE,
                            check.names = FALSE)
knitr::kable(expected)
```

## Reports you can hand to somebody

`render_results()` writes the already-computed analysis results. It does not
rerun the plan. `render_report()` builds the complete research report, including
the codebook, screening, measurement sections, analysis, models, and
reproducibility metadata.

```{r report}
# The report is rendered only when at least one analysis succeeded.
if (!is.null(attr(results, "status")) && attr(results, "status")$succeeded > 0) {
  render_results(results, instrument, output_file = "analysis-results.html")
  render_report(instrument, responses, output_file = "full-report.html",
                show_code = TRUE)
} else {
  cat("Every analysis failed, so no reports were rendered. See the errors above.\n")
}
```

## Checking these numbers against other software

Every demo ships its data, a codebook of variable and value labels, and the
results surveyframe produced. Take them into `psych`, SPSS, JASP, jamovi or
Stata and compare.

```{r export}
#| eval: false
sframe_export_labelled(responses, instrument, "{{DEMO_NAME}}.sav")
```

The `.sav` arrives with the question wording and the response options already
attached, so your variables read as questions rather than as codes.

If a number comes out differently, please
[open an issue](https://github.com/MohammedAliSharafuddin/surveyframe/issues)
with the demo name, the software you used, and both results.

## Adapt this for your own survey

1. Replace `sframe_demo("{{DEMO_NAME}}")` with your own `read_sframe()` call,
   or build an instrument with `sf_instrument()`.
2. Declare your analysis plan **before** you collect anything.
3. Everything below stays as it is.
