---
title: "{{TITLE}}"
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(surveyframe)

`%||%` <- function(x, y) if (is.null(x)) y else x

show_value <- function(x) {
  if (is.null(x) || length(x) == 0) return("")
  if (is.list(x)) {
    return(paste(sprintf("%s: %s", names(x),
                         vapply(x, show_value, character(1))), collapse = "; "))
  }
  paste(as.character(x), collapse = ", ")
}

plan_table <- function(instrument) {
  plan <- sf_plan(instrument)
  if (!length(plan)) return(data.frame())
  do.call(rbind, lapply(seq_along(plan), function(i) {
    b <- plan[[i]]
    data.frame(
      order = i,
      id = b$id,
      question = b$research_question,
      family = b$family %||% "",
      method = b$method %||% b$test %||% "",
      roles = show_value(b$roles %||% b$variables),
      options = show_value(b$options),
      decision_rule = b$decision_rule %||% b$interpretation %||% "",
      needs_data = !isFALSE(b$requires_data),
      stringsAsFactors = FALSE,
      check.names = FALSE
    )
  }))
}
```

## The study

This notebook runs your instrument end to end: load the instrument and its
responses, run the pre-declared analysis plan, and render a report. It reads
`{{SFRAME_FILE}}` and `{{CSV_FILE}}` from the same folder as this notebook,
so keep the 3 files together.

```{r load}
instrument <- read_sframe("{{SFRAME_FILE}}")
# strict = FALSE: the CSV may carry extra columns (respondent_id,
# submitted_at, and similar) that aren't declared in the instrument itself.
responses  <- read_responses("{{CSV_FILE}}", instrument, strict = FALSE)
instrument
```

## The questions

```{r items}
codebook <- as.data.frame(codebook_report(instrument))
knitr::kable(codebook[, c("id", "label", "type")])
```

## The plan, declared before the data existed

This is what makes the analysis a contract rather than a search. The blocks
below were written into the instrument at design time.

```{r plan}
knitr::kable(plan_table(instrument))
```

The first column is both the run and report sequence. `run_analysis_plan()`
executes the plan from top to bottom. In SurveyBuilder, drag plan cards on the
Plan or Run Preview tab before examining results. In R, reorder by stable block
ID, validate, and save before data collection:

```{r reorder, eval = FALSE}
plan <- sf_plan(instrument)
ids <- vapply(plan, `[[`, character(1), "id")
wanted_order <- rev(ids)             # replace with the order you planned
stopifnot(setequal(wanted_order, ids))
sf_plan(instrument) <- plan[match(wanted_order, ids)]
instrument <- as_sframe(validate_sframe(instrument, strict = TRUE))
```

A change after collection begins should retain the earlier instrument and be
disclosed with `amend_sframe()`.

## Validate and screen the inputs

```{r validate}
validation <- validate_sframe(instrument, strict = FALSE)
validation
dim(responses)
head(responses, 3)
```

```{r screen}
missing <- missing_data_report(responses, instrument)
knitr::kable(as.data.frame(missing))

quality <- quality_report(responses, instrument)
knitr::kable(as.data.frame(quality))
```

```{r reliability}
if (length(sf_scales(instrument)) > 0 &&
    requireNamespace("psych", quietly = TRUE)) {
  reliability <- reliability_report(
    responses, instrument,
    omega = TRUE
  )
  knitr::kable(as.data.frame(reliability))
} else if (length(sf_scales(instrument)) == 0) {
  cat("This instrument has no scored scales, so reliability is not applicable.\n")
} else {
  cat("Install the optional psych package to compute reliability.\n")
}

## Run the plan once

```{r run}
results <- run_analysis_plan(
  responses,
  instrument,
  plots = requireNamespace("ggplot2", quietly = TRUE),
  seed = 20260828L,
  strict = FALSE
)
status <- attr(results, "status")
status
attr(results, "seed")
knitr::kable(as.data.frame(results))
```

A normal non-strict return may contain failed blocks, so inspect `status`. For
a verification or release run, use `strict = TRUE` to stop on any failed block.

```{r strict-run, eval = FALSE}
results <- run_analysis_plan(
  responses, instrument, plots = TRUE, seed = 20260828L, strict = TRUE
)
```

## Read the complete results

```{r result-details, results = "asis"}
plan <- sf_plan(instrument)
for (b in results) {
  planned <- plan[[match(b$block_id, vapply(plan, `[[`, character(1), "id"))]]
  cat("\n###", b$block_id, b$research_question, "\n\n")
  cat("**Planned method:**", planned$method %||% planned$test, "\n\n")
  cat("**Roles:**", show_value(planned$roles %||% planned$variables), "\n\n")
  if (nzchar(planned$decision_rule %||% planned$interpretation %||% "")) {
    cat("**Planned decision rule:**",
        planned$decision_rule %||% planned$interpretation, "\n\n")
  }
  if (!is.null(b$error)) {
    cat("**This analysis failed.** ", b$error, "\n\n", sep = "")
    next
  }
  if (!is.null(b$apa)) cat(b$apa, "\n\n")
  if (!is.null(b$effect_label)) cat("**Effect:**", b$effect_label, "\n\n")
  if (is.data.frame(b$table)) print(knitr::kable(b$table))
  if (!is.null(b$syntax)) cat("\n```r\n", b$syntax, "\n```\n", sep = "")
  if (!is.null(b$plot)) print(b$plot)
  if (length(b$diagnostic_plots %||% list())) {
    for (p in b$diagnostic_plots) print(p)
  }
  resolved <- analysis_syntax(b)
  if (length(resolved)) cat("\n**Resolved R call**\n\n```r\n",
                            paste(resolved, collapse = "\n"), "\n```\n", sep = "")
  if (nzchar(b$prompt %||% "")) cat("\n**Interpretation prompt:**", b$prompt, "\n\n")
  refs <- unlist(b$citations %||% character(0), use.names = FALSE)
  if (length(refs)) cat("**References:**", paste(refs, collapse = "; "), "\n\n")
}
if (!is.null(status) && status$failed > 0) {
  cat("\n", status$failed, " of ", status$blocks, " analyses failed: ",
      paste(status$failed_blocks, collapse = ", "), ".\n", sep = "")
}
```

## Reports you can hand to somebody

`render_results()` writes the results already computed above. `render_report()`
builds the complete report, including the codebook, screening, measurement,
analysis, models, and reproducibility metadata.

```{r report}
# The report is rendered only when at least one analysis succeeded.
if (!is.null(attr(results, "status")) && attr(results, "status")$succeeded > 0) {
  render_results(results, instrument, output_file = "analysis-results.html")
  render_report(instrument, responses, output_file = "full-report.html",
                show_code = TRUE)
} else {
  cat("Every analysis failed, so no reports were rendered. See the errors above.\n")
}
```

## Checking these numbers against other software

```{r export}
#| eval: false
sframe_export_labelled(responses, instrument, "responses.sav")
```

The `.sav` arrives with the question wording and the response options already
attached, so your variables read as questions rather than as codes. Take it
into `psych`, SPSS, JASP, jamovi or Stata and compare.
