---
lang: en-GB
title: "Multi-criteria decision analysis"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Multi-criteria decision analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(surveyframe)
library(knitr)
set.seed(2026)

has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(fig.width = 7, fig.height = 4.2, dpi = 96)
```

```{=html}
<style>
body { color: #1a1a2e; }
h1, h2, h3 { color: #1a1a2e; }
a { color: #0e7c7a; }
table { border-collapse: collapse; margin: 1em 0; }
table caption { caption-side: top; font-style: italic; color: #444; padding-bottom: .3em; }
th { border-top: 2px solid #1a1a2e; border-bottom: 1px solid #1a1a2e; padding: 6px 12px; }
td { padding: 5px 12px; border: none; }
tbody tr:last-child td { border-bottom: 2px solid #1a1a2e; }
/* WCAG 2.2 AA pass: darker link and syntax-token colours (at least
   4.5:1 on the #f7f7f7 code background), wrapped code lines instead of a
   keyboard-inaccessible scroll region, empty per-line anchors removed from
   the accessibility tree, and 24px minimum TOC link targets. */
code span.at { color: #576419; }
code span.dv, code span.fl, code span.bn { color: #276245; }
code span.co { color: #396a80; }
pre, pre code { white-space: pre-wrap; word-break: break-word; }
div.sourceCode { overflow: visible; }
pre.sourceCode a:empty { display: none; }
#TOC a { display: inline-block; min-height: 24px; }
</style>
```

## What this vignette covers

Multi-criteria decision analysis ranks a handful of alternatives against
several criteria that pull in different directions. surveyframe treats it the
same way it treats any other analysis: the method, its inputs, and the roles
they play are declared in the instrument before data is collected, and the
analysis is the execution of that declaration.

The worked example is a hotel choosing between 5 suppliers on 4 criteria. It
ships with the package, so every number below is reproducible.

```{r load}
demo <- read_sframe(system.file("extdata", "hotel_supplier_mcdm.sframe",
                                package = "surveyframe"))
responses <- utils::read.csv(
  system.file("extdata", "hotel_supplier_mcdm_responses.csv",
              package = "surveyframe"),
  stringsAsFactors = FALSE
)

c(respondents = nrow(responses), criteria = 4, suppliers = 5)
```

## Where the numbers come from

An MCDM result needs two things: a performance matrix saying how each
alternative scores on each criterion, and a weight vector saying how much
each criterion matters. They can come from different places, and being
explicit about which is the difference between a defensible result and a
plausible-looking one.

This instrument declares three sources at once, so the example exercises all
of them.

```{r sources, echo = FALSE}
kable(
  data.frame(
    Source = c("Pairwise comparison", "Constant sum", "Rated matrix",
               "Researcher supplied"),
    Item = c("crit_pairs", "crit_points",
             "rate_service and 3 more", "declared in the plan block"),
    Provides = c("Criterion weights", "Criterion weights",
                 "Performance matrix", "Performance matrix"),
    How = c("Respondents judge each pair on the Saaty 1 to 9 scale",
            "Respondents divide 100 points across the criteria",
            "Respondents rate every supplier on every criterion",
            "Audited figures the researcher enters directly"),
    check.names = FALSE, stringsAsFactors = FALSE
  ),
  row.names = FALSE, align = c("l", "l", "l", "l"),
  caption = "The 4 declared input sources in the worked example."
)
```

Each result records which source it used, so a report can say where its
numbers came from rather than leaving a reader to assume.

## Criterion weights from pairwise judgements

The first block asks what weight each criterion carries. Respondents compared
every pair of criteria, and AHP turns those judgements into weights.

```{r ahp}
results <- run_analysis_plan(responses, demo, plots = has_ggplot)
ahp <- results[["RQ1"]]

kable(ahp$table, row.names = FALSE,
      caption = "Criterion weights derived from pairwise judgements.")
```

Pairwise judgements can contradict each other. If service beats price, and
price beats delivery, then service ought to beat delivery by roughly the
product of the two. The consistency ratio measures how far the judgements
depart from that.

```{r consistency}
round(ahp$cr, 4)
```

Saaty's convention treats a consistency ratio below 0.10 as acceptable. This
one is well inside it. A ratio above 0.10 should be reported alongside the
result, since it still yields a usable weight set. `options$cr_filter = TRUE`
will drop individual respondents above the threshold before aggregation if a
study has pre-declared that rule.

```{r ahp-plot, fig.alt = "Bar chart of the four criterion weights derived from pairwise comparisons, with service carrying the largest weight and delivery the smallest.", eval = has_ggplot}
ahp$plot
```

## Ranking suppliers on audited figures

The second block ranks the suppliers using a performance matrix the
researcher supplies, combined with the weights the respondents produced. This
is the common hybrid: measured facts about the alternatives, weighted by the
people who will live with the decision.

```{r topsis-supplied}
audited <- results[["RQ2"]]

kable(audited$table, row.names = FALSE,
      caption = paste("TOPSIS ranking on audited figures, weighted by",
                      "collected judgements."))

c(weights = audited$weights_source, matrix = audited$matrix_source)
```

Note the `criteria_types` this block declares: service and location are
`benefit` criteria where more is better, while price and delivery time are
`cost` criteria where less is better. Getting that wrong silently inverts the
ranking, which is why it is declared in the instrument rather than inferred.

## Ranking suppliers on collected ratings

The third block answers a different question with the same method: not which
supplier the audited figures favour, but which one the staff rate best. The
performance matrix is built from respondents' ratings, and the weights come
from the constant-sum question instead of the pairwise one.

```{r topsis-rated}
rated <- results[["RQ3"]]

kable(rated$table, row.names = FALSE,
      caption = paste("TOPSIS ranking on staff ratings, weighted by the",
                      "constant-sum question."))
```

The two rankings do not agree, and that is the useful part. Comparing them is
a finding rather than a problem to be resolved.

```{r compare, echo = FALSE}
cmp <- merge(
  audited$table[, c("Alternative", "Rank")],
  rated$table[, c("Alternative", "Rank")],
  by = "Alternative", suffixes = c("_audited", "_rated")
)
cmp <- cmp[order(cmp$Rank_audited), ]
kable(cmp, row.names = FALSE,
      col.names = c("Supplier", "Rank on audited figures",
                    "Rank on staff ratings"),
      caption = "The same method, two declared input sources, two answers.")
```

### A trap worth naming

All 4 criteria in the rated block are declared `benefit`, including price.
That is correct here only because the question asked about **value for money**,
where a higher rating is better. Had it asked respondents to rate price
directly, a higher rating would mean more expensive and the criterion would be
a `cost`.

Nothing in the data distinguishes those two cases. The wording of the question
does, and the declaration has to match it. This is the single easiest way to
produce a confident, precise, and completely inverted ranking.

## How much do the weights matter?

A ranking produced from collected weights inherits their uncertainty. Before
reporting a winner, it is worth asking how much of that result survives a
small change in the weights.

```{r sensitivity}
sens <- sensitivity_analysis(
  x = matrix(c(4.1, 3.0, 210, 36,
               3.6, 4.5, 180, 48,
               4.8, 2.5, 260, 24,
               3.9, 4.0, 150, 72,
               4.4, 3.8, 230, 30),
             nrow = 5, byrow = TRUE),
  weights        = audited$weights,
  criteria_types = c("benefit", "benefit", "cost", "cost"),
  method         = "topsis",
  alternatives   = c("Alpha", "Basilica", "Coral", "Dhoni", "Equator"),
  criteria       = c("service", "location", "price", "delivery")
)

sens
```

Each criterion's weight is nudged up and down by 5 percent, renormalised, and
the ranking is recomputed. `rho` is the rank correlation with the original
ranking, and `top_changed` records whether the leading alternative changed.

```{r sensitivity-table}
kable(as.data.frame(sens), row.names = FALSE,
      caption = "Ranking stability under a 5 percent change in each weight.")
```

```{r sensitivity-plot, fig.alt = "Bar chart of rank correlation for each criterion perturbed up and down, against a dashed reference line at one marking an unchanged ranking.", eval = has_ggplot}
plot(sens)
```

This example is worth reading closely, because it is not the clean case. Four
of the 8 perturbations changed the ranking, so `stable` is `FALSE`. But
`top_changed` is `FALSE` throughout: the order shuffles among the middle
suppliers while Equator stays first under every nudge.

That distinction is the whole point of running this. "Equator ranks first, and
that holds under a 5 percent change in any single weight" is a defensible
claim. "The ranking is Equator, Coral, Basilica, Alpha, Dhoni" is not, because
positions 2 to 5 move. Reporting the full ranking as though it were as solid
as the winner would overstate what the data supports.

A result where `top_changed` is `TRUE` anywhere deserves a stronger caveat
still, and one where `stable` is `TRUE` throughout can be reported as robust
to the weights.

## Which criteria drive the others

The criteria are not independent. Delivery speed and price move together, and
service quality may drive both. DEMATEL asks respondents how strongly each
factor influences each other factor and separates the causes from the effects.

```{r dematel}
dematel <- results[["RQ4"]]

kable(dematel$table, row.names = FALSE,
      caption = "DEMATEL cause and effect classification.")
```

The prominence column measures how involved a criterion is in the system
overall, and the relation column separates the drivers from the driven. A
criterion with a positive relation value influences others more than it is
influenced.

```{r dematel-plot, fig.alt = "Influence map plotting prominence against relation for the four criteria, separating causal drivers above the axis from affected criteria below it.", eval = has_ggplot}
dematel$plot
```

Note that the influence question uses a different scale from the AHP one. AHP
reads reciprocal relative importance on Saaty's 1 to 9 ratio scale, while
DEMATEL reads directed 0 to 4 influence with no reciprocity. They are not
interchangeable, and surveyframe refuses to pair one with the other's method
at validation time rather than returning plausible numbers from meaningless
input.

## Reporting the whole plan

Because every block is declared in the instrument, the whole analysis is one
call and the report writes itself in the same order the plan was declared.

```{r summary, echo = FALSE}
# as.data.frame() flattens the results to one row per block, so the whole
# summary is a column selection rather than a loop over internals.
results_df <- as.data.frame(results)
summary_df <- data.frame(
  RQ = results_df$block_id,
  Question = results_df$research_question,
  Method = toupper(results_df$test),
  Result = results_df$apa,
  check.names = FALSE, stringsAsFactors = FALSE
)
kable(summary_df, row.names = FALSE,
      col.names = c("RQ", "Research question", "Method", "Result"),
      caption = "The declared analysis plan and what each block returned.")
```

## What surveyframe does not do here

The decision family ranks and weights. It does not tell a researcher which
method to use, and the choice matters: the 10 available methods encode
different assumptions about how criteria trade off against one another.

Two limits are worth stating plainly. surveyframe does not estimate choice
models, so `sf_conjoint_design()` declares a conjoint design without
analysing its responses. And PROMETHEE defaults to Brans and Vincke's type I
step function rather than the linear function some implementations default
to, because the linear function needs thresholds that are commonly derived
from the data range, which makes a result depend on a choice nobody declared.
See `?sframe_decision_options` for the detail, including how far rankings
move between the two.
